Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-27 parallel-tractability-J-v9/response.tex

\begin{quote}
Finally, the algorithms assume each processor handling a rule instantiation. I wonder how it is implemented in ParallelDHL.
\end{quote}

...

\begin{quote}
I list some relatively minor issues regarding presentation and technical details:

- p20, it mentions a group of materialisation rules in YAGO which belong to $D_{\text{dhl}}$. It deserves a clarification on what these rules are like and why they do not fall into existing parallel tractable classes.\\

- p20, Table 2, it should mention how these ontologies were selected. \\

- p21, the discussion on line graph lg1 is confusing (i.e.\ ``line-2 stays higher than line-6, but lower than line-4''). Maybe I was looking at the wrong place. \\

- p25, the section heading ``Reference'' occur twice.
\end{quote}

### 2018-01-27 parallel-tractability-J-v9/parallel-tractability-J.tex
fixed minor issues

\begin{abstract}
Materialization is an important reasoning service for applications built
on the ontology languages, such as OWL. With a
rapid growth of semantic data, it is challenging to perform
materialization on large-scale ontologies efficiently.
Current work proposes parallel algorithms and employs computing
platforms of high-performance to make materialization sufficiently efficient and scalable in practice.
However, there do exist ontologies, for which
materialization cannot always be improved by parallelism (we use experiments to show it in Section~\ref{sec:evaluation}).
On the other hand, some well-known large-scale ontologies, such as YAGO, have been shown to have
good performance for parallel reasoning, but they are expressed in ontology languages
that are not parallelly tractable in theory, i.e., the efficiency of reasoning may not be improved
on a parallel implementation.
This motivates us to study the problem of \emph{parallel tractability} of ontology
materialization from a theoretical perspective.
That is, we aim to identify the ontologies for which materialization is
parallelly tractable, i.e., in \texttt{NC} complexity.
The results of the parallel tractability can be further used to optimize
materialization algorithms and to guide ontology engineers in creating parallelly tractable ontologies.

@@

There has been work that proposes parallel reasoning algorithms and employs parallel computing platforms
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balance workload,
data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, in order to make reasoning tasks more efficient
in parallel, several work utilizes different kinds of high-performance computing platforms, like supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that, some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, we show in Section~\ref{sec:evaluation} that, there do exist some ontologies, for which materialization cannot
always be improved by parallelism. The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is a free variable.
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

@@

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent node of node $v_2$ and node $v_2$ is the child node
of node $v_1$. Further, $\mathcal{G}$ satisfies the following condition:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $\forall H,B_1,...,B_n\in V$ such that $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$, we have that
  $B_1,...,B_n\rightarrow H\in P^*$.
\end{enumerate}
\end{definition}

@@

\noindent\texttt{Algorithm~$\mathsf{PRC}$}. This algorithm has inputs (1)
a DHL$(\circ)$ ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$
and its datalog program $P=\langle R, \textbf{I}\rangle$;
(2) a (partial) materialization graph $\mathcal{G}$ that is constructed from $P$.
This algorithm outputs a \texttt{rch} relation $S_{\textit{\tiny rch}}$ that
is computed from the following rule instantiations:

\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
\item $\forall A(a)\rightarrow B(a)$ where $\mathcal{O}\models A\sqsubseteq_* B$ and $A(a)\in\mathcal{G}$,
    add \texttt{rch}$(A(a),B(a))$ to $S_{\textit{\tiny rch}}$;

\item $\forall R(a,b),A(b)\rightarrow B(a)$ where $\exists R.A\sqsubseteq B\in\mathcal{T}$
    and $R(a,b)\in\mathcal{G}$, add \texttt{rch}$(A(b),B(a))$ to $S_{\textit{\tiny rch}}$;

\item $\forall A_1(a),A_2(a)\rightarrow B(a)$ where $A_1\sqcap A_2\sqsubseteq B\in\mathcal{T}$,
    if $A_1(a)\in\mathcal{G}$, add \texttt{rch}$(A_2(a),B(a))$ to $S_{\textit{\tiny rch}}$;
    if $A_2(a)\in\mathcal{G}$, add \texttt{rch}$(A_1(a),B(a))$ to $S_{\textit{\tiny rch}}$;

\item $\forall R(a,b)\rightarrow S(a,b)$ where $\mathcal{O}\models R\sqsubseteq_* S$ and $R(a,b)\in\mathcal{G}$,
    add \texttt{rch}$(R(a,b)$, $S(a,b))$ to $S_{\textit{\tiny rch}}$;

\item $\forall R(a,b)\rightarrow S(b,a)$ where $\mathcal{O}\models R\sqsubseteq_* S^-$
    and $R(a,b)\in\mathcal{G}$, add \texttt{rch}$(R(a,b)$, $S(b,a))$ to $S_{\textit{\tiny rch}}$;

\item $\forall R_1(a,b),R_2(b,c)\rightarrow R_3(a,c)$ where $R_1\circ R_2\sqsubseteq R_3$ (the case
    where $R_1\equiv R_2\equiv R_3$ is also involved),
    if $R_1(a,b)\in\mathcal{G}$, add \texttt{rch}$(R_2(b,c),R_3(a,c))$ to $S_{\textit{\tiny rch}}$;
    If $R_2(b,c)\in\mathcal{G}$, add \texttt{rch}$(R_1(a,b)$, $R_3(a,c))$ to $S_{\textit{\tiny rch}}$.\hfill$\Box$
\end{enumerate}

@@

We further process the collected data and fill the results in Table~3, where each cell corresponds
to a test ontology (see the row label) and a new generated ABox (distinguished by the column labels);
the three values from above to below in each cell are: (1) \emph{the minimal time ratio} - the minimal reasoning time
of ParallelDHL devided by the minimal reasoning time of RDFox;
(2) \emph{the average speedup of ParallelDHL};\footnote{Suppose $T_i$ is the reasoning time with $i$ threads allocated,
the average speedup is $\frac{1}{4}(\frac{T_1}{T_2}+\frac{T_2}{T_4}+\frac{T_4}{T_6}+\frac{T_6}{T_8})$.}
(3) \emph{the average speedup of RDFox}.
The minimal time ratio describes the performance of ParallelDHL by using RDFox as the baseline.
The indicator speedup and its derived indicators are the most common tools used for measuring the capacity of parallelism \cite{MotikNPHO14,KazakovKS14,UrbaniKMHB12}.
Here, we use the average speedup \cite{ichiyoshiK92} to describe the average capacity of parallelism with different
threads allocated.
In the following, we give the detailed analysis based on the contents in Figure~\ref{fig:linegraph} and Table~3.

>>>>

\begin{abstract}
Materialization is an important reasoning service for applications built
on the ontology languages, such as OWL. With a
rapid growth of semantic data, it is challenging to perform
materialization on large-scale ontologies efficiently.
Current work proposes parallel algorithms and employs computing
platforms of high-performance to make materialization sufficiently efficient and scalable in practice.
However, there do exist ontologies, for which
materialization cannot always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}).
On the other hand, some well-known large-scale ontologies, such as YAGO, have been shown to have
good performance for parallel reasoning, but they are expressed in ontology languages
that are not parallelly tractable in theory, i.e., the efficiency of reasoning may not be improved
on a parallel implementation.
This motivates us to study the problem of \emph{parallel tractability} of ontology
materialization from a theoretical perspective.
That is, we aim to identify the ontologies for which materialization is
parallelly tractable, i.e., in \texttt{NC} complexity.
The results of the parallel tractability can be further used to optimize
materialization algorithms and to guide ontology engineers in creating parallelly tractable ontologies.

@@

There has been work proposing parallel reasoning algorithms and employing parallel computing platforms
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balancing workload
and data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, several work utilizes different kinds of computing platforms to make reasoning tasks more efficient in parallel, such as supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, there do exist ontologies for which materialization cannot
always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}). 
The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is called a \emph{free variable} (it is
not occurring in the head atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

@@

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent (node) of node $v_2$ and node $v_2$ is the child (node)
of node $v_1$. Further, $\forall H,B_1,...,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item $H$ is the original fact of $P$ if $H$ has the in-degree of $0$;
  \item $B_1,...,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

@@

\noindent\texttt{Algorithm~$\mathsf{PRC}$}. This algorithm has inputs (1)
a DHL$(\circ)$ ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$
and its datalog program $P=\langle R, \textbf{I}\rangle$;
(2) a (partial) materialization graph $\mathcal{G}=\langle V,E\rangle$ that is constructed from $P$.
This algorithm outputs a \texttt{rch} relation $S_{\textit{\tiny rch}}$ that
is computed as follows:

\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
\item add \texttt{rch}$(A(a),B(a))$ to $S_{\textit{\tiny rch}}$ where $A(a)\rightarrow B(a)\in P^*$, $\mathcal{O}\models A\sqsubseteq_* B$ and $A(a)\in V$;

\item add \texttt{rch}$(A(b),B(a))$ to $S_{\textit{\tiny rch}}$ where $R(a,b),A(b)\rightarrow B(a)\in P^*$, $\exists R.A\sqsubseteq B\in\mathcal{T}$ and $R(a,b)\in V$;

\item add \texttt{rch}$(A_2(a),B(a))$ to $S_{\textit{\tiny rch}}$ where $A_1(a),A_2(a)\rightarrow B(a)\in P^*$,
    $A_1\sqcap A_2\sqsubseteq B\in\mathcal{T}$ and $A_1(a)\in V$;

\item add \texttt{rch}$(A_1(a),B(a))$ to $S_{\textit{\tiny rch}}$ where $A_1(a),A_2(a)\rightarrow B(a)\in P^*$,
    $A_1\sqcap A_2\sqsubseteq B\in\mathcal{T}$ and $A_2(a)\in V$;

\item add \texttt{rch}$(R(a,b)$, $S(a,b))$ to $S_{\textit{\tiny rch}}$ where $R(a,b)\rightarrow S(a,b)\in P^*$,
    $\mathcal{O}\models R\sqsubseteq_* S$ and $R(a,b)\in V$;

\item add \texttt{rch}$(R(a,b)$, $S(b,a))$ to $S_{\textit{\tiny rch}}$ where $R(a,b)\rightarrow S(b,a)\in P^*$,
    $\mathcal{O}\models R\sqsubseteq_* S^-$ and $R(a,b)\in V$;

\item add \texttt{rch}$(R_2(b,c),R_3(a,c))$ to $S_{\textit{\tiny rch}}$ where $R_1(a,b),R_2(b,c)\rightarrow R_3(a,c)\in P^*$, 
    $R_1\circ R_2\sqsubseteq R_3\in\mathcal{R}$ (the case
    where $R_1\equiv R_2\equiv R_3$ is also involved) and $R_1(a,b)\in V$;

\item add \texttt{rch}$(R_1(a,b)$, $R_3(a,c))$ to $S_{\textit{\tiny rch}}$ where $R_1(a,b),R_2(b,c)\rightarrow R_3(a,c)\in P^*$,
    $R_1\circ R_2\sqsubseteq R_3\in\mathcal{R}$ and $R_2(b,c)\in V$.\hfill$\Box$
\end{enumerate}

@@

We further process the collected data and fill the results in Table~3, where each cell corresponds
to a test ontology (see the row label) and a new generated ABox (distinguished by the column labels);
the three values from above to below in each cell are: (1) \emph{the minimal time ratio} - the minimal reasoning time
of ParallelDHL divided by the minimal reasoning time of RDFox;
(2) \emph{the average speedup of ParallelDHL};\footnote{Suppose $T_i$ is the reasoning time with $i$ threads allocated,
the average speedup is $\frac{1}{4}(\frac{T_1}{T_2}+\frac{T_2}{T_4}+\frac{T_4}{T_6}+\frac{T_6}{T_8})$.}
(3) \emph{the average speedup of RDFox}.
The minimal time ratio describes the performance of ParallelDHL by using RDFox as the baseline.
The indicator speedup and its derived indicators are the most common tools used for measuring the capacity of parallelism \cite{MotikNPHO14,KazakovKS14,UrbaniKMHB12}.
Here, we use the average speedup \cite{ichiyoshiK92} to describe the average capacity of parallelism with different
threads allocated.
In the following, we give the detailed analysis based on the contents in Figure~\ref{fig:linegraph} and Table~3.
